---
title: "Graphics with ggplot2"
author: "Alok K. Shukla"
date: "12/11/2016"
output:
  html_document:
    highlight: tango
    theme: flatly
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: tango
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

Sample dataset.

```{r,warning=FALSE,message=FALSE}
housing <- read.csv("dataSets/landdata-states.csv")
head(housing[1:5])
```


`ggplot2` VS Base for simple graphs
-----------------------------------

Base graphics histogram example:

```{r,warning=FALSE,message=FALSE}
  hist(housing$Home.Value)
```

`ggplot2` histogram example:

```{r,warning=FALSE,message=FALSE}
  ggplot(housing, aes(x = Home.Value)) +
    geom_histogram()
```
`ggplot2` Base graphics VS `ggplot` for more complex graphs:
------------------------------------------------------------

Base colored scatter plot example:

```{r,warning=FALSE,message=FALSE}
  plot(Home.Value ~ Date,
       data=subset(housing, State == "MA"))
  points(Home.Value ~ Date, col="red",
         data=subset(housing, State == "TX"))
  legend(19750, 400000,
         c("MA", "TX"), title="State",
         col=c("black", "red"),
         pch=c(1, 1))
```

`ggplot2` colored scatter plot example:

```{r,warning=FALSE,message=FALSE}
  ggplot(subset(housing, State %in% c("MA", "TX")),
         aes(x=Date,
             y=Home.Value,
             color=State))+
    geom_point()
```

Geometric Objects And Aesthetics
================================

Aesthetic Mapping
-----------------

In ggplot land *aesthetic* means "something you can see". Examples include:

-   position (i.e., on the x and y axes)
-   color ("outside" color)
-   fill ("inside" color)
-   shape (of points)
-   linetype
-   size

Aesthetic mappings are set with the `aes()` function.

Geometic Objects (`geom`)
-------------------------

Geometric objects are the actual marks we put on a plot. Examples include:

-   points (`geom_point`, for scatter plots, dot plots, etc)
-   lines (`geom_line`, for time series, trend lines, etc)
-   boxplot (`geom_boxplot`, for, well, boxplots!)

A plot must have at least one geom; there is no upper limit. You can add a geom to a plot using the `+` operator.

Points (Scatterplot)
--------------------

Now that we know about geometric objects and aesthetic mapping, we can make a ggplot. `geom_point` requires mappings for x and y, all others are optional.

```{r,warning=FALSE,message=FALSE}
  hp2001Q1 <- subset(housing, Date == 20011) 
  ggplot(hp2001Q1,
         aes(y = Structure.Cost, x = Land.Value)) +
    geom_point()
```

Lines (Prediction Line)
-----------------------

A plot constructed with `ggplot` can have more than one geom. In that case the mappings established in the `ggplot()` call are plot defaults that can be added to or overridden. Our plot could use a regression line:

```{r,warning=FALSE,message=FALSE}
  hp2001Q1$pred.SC <- predict(lm(Structure.Cost ~ Land.Value, data = hp2001Q1))

  ## Added mapping
  p1 <- ggplot(hp2001Q1, aes(x = Land.Value, y = Structure.Cost))
  ## Overridden
  p1 + geom_point(aes(color = Home.Value)) +
    geom_line(aes(y = pred.SC))
```
Smoothers
---------

Not all geometric objects are simple shapes--the smooth geom includes a line and a ribbon.

```{r,warning=FALSE,message=FALSE}
  p1 +
    geom_point(aes(color = Home.Value)) +
    geom_smooth()
```

Text (Label Points)
-------------------

Each `geom` accepts a particualar set of mappings--for example `geom_text()` accepts a `labels` mapping.

```{r,warning=FALSE,message=FALSE}
  p1 + 
    geom_text(aes(label=State), size = 2)
```

```{r,warning=FALSE,message=FALSE}
  ##install.packages("ggrepel") 
  library("ggrepel")
  p1 + 
    geom_point() + 
    geom_text_repel(aes(label=State), size = 3)
```

Aesthetic Mapping VS Assignment
-------------------------------

Note that variables are mapped to aesthetics with the `aes()` function, while fixed aesthetics are set outside the `aes()` call. This sometimes leads to confusion, as in this example:

```{r,warning=FALSE,message=FALSE}
## Incorrect
## p1 + geom_point(aes(size = 2)
## Correct    
  p1 + geom_point(color="red") # this is fine -- all points red
```


Mapping Variables To Other Aesthetics
-------------------------------------

Other aesthetics are mapped in the same way as x and y in the previous example.

```{r,warning=FALSE,message=FALSE}
  p1 + geom_point(aes(color=Home.Value, shape = region))
```

Exercise I
----------

The data for the exercises is available in the `dataSets/EconomistData.csv` file. Read it in with

```{r,warning=FALSE,message=FALSE}
  dat <- read.csv("dataSets/EconomistData.csv")
```


These data consist of *Human Development Index* and *Corruption Perception Index* scores for several countries.

1.  Create a scatter plot with CPI on the x axis and HDI on the y axis.
2.  Color the points in the previous plot blue.
3.  Color the points in the previous plot according to *Region*.
4.  Create boxplots of CPI by Region
5.  Overlay points on top of the box plots

Solution
---------
1. Create a scatter plot with CPI on the x axis and HDI on the y axis.

```{r,warning=FALSE,message=FALSE}
p1 = ggplot(dat, aes(x = CPI, y = HDI)) 
p1 + geom_point()
```
2.  Color the points in the previous plot blue.

```{r,warning=FALSE,message=FALSE}
p1 +  geom_point(color = "blue")
```
3.  Color the points in the previous plot according to *Region*.

```{r,warning=FALSE,message=FALSE}
p1+ geom_point(aes(color = Region))
```

4.  Create boxplots of CPI by Region

```{r,warning=FALSE,message=FALSE}
ggplot(dat, aes(x = Region, y = CPI)) + geom_boxplot()
```

5.  Overlay points on top of the box plots

```{r,warning=FALSE,message=FALSE}
ggplot(dat, aes(x = Region, y = CPI)) + geom_boxplot() +geom_point()
```


Statistical Transformations
===========================

Statistical Transformations
---------------------------

Some plot types (such as scatterplots) do not require transformations--each point is plotted at x and y coordinates equal to the original value. Other plots, such as boxplots, histograms, prediction lines etc. require statistical transformations:

-   for a boxplot the y values must be transformed to the median and 1.5(IQR)
-   for a smoother smother the y values must be transformed into predicted values

Each `geom` has a default statistic, but these can be changed. For example, the default statistic for `geom_bar` is `stat_count`:

```{r,warning=FALSE,message=FALSE}
  args(geom_histogram)
  args(stat_bin)
```

Setting Statistical Transformation Arguments
--------------------------------------------

Arguments to `stat_` functions can be passed through `geom_` functions. This can be slightly annoying because in order to change it you have to first determine which stat the geom uses, then determine the arguments to that stat.

For example, here is the default histogram of Home.Value:

```{r,warning=FALSE,message=FALSE}
  p2 <- ggplot(housing, aes(x = Home.Value))
  p2 + geom_histogram()
```
The binwidth looks reasonable by default, but we can change it by passing the `binwidth` argument to the `stat_bin` function:

```{r,warning=FALSE,message=FALSE}
  p2 + geom_histogram(stat = "bin", binwidth=4000)
```

Changing The Statistical Transformation
---------------------------------------

Sometimes the default statistical transformation is not what you need. This is often the case with pre-summarized data:

```{r,warning=FALSE,message=FALSE}
  housing.sum <- aggregate(housing["Home.Value"], housing["State"], FUN=mean)
  rbind(head(housing.sum), tail(housing.sum))
```

```{r,warning=FALSE,message=FALSE}
## Already transformed
## ggplot(housing.sum, aes(x=State, y=Home.Value)) + geom_bar()
```


What is the problem with the previous plot? Basically we take binned and summarized data and ask ggplot to bin and summarize it again (remember, `geom_bar` defaults to `stat ` stat<sub>count</sub>=); obviously this will not work. We can fix it by telling `geom_bar` to use a different statistical transformation function:

```{r,warning=FALSE,message=FALSE}
  ggplot(housing.sum, aes(x=State, y=Home.Value)) + geom_bar(stat="identity")
```